---
title: "Baseball"
author: "Bedros, Haig; Dela Cruz, Sandra; Hugh, Tiffany; Li, Yanyi"
date: "2025-02-12"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
#Install package sand import dataset
install.packages(c("tidyverse", "knitr", "readr","DataExplorer"))
library(tidyverse)
#library(knitr)
library(DataExplorer)
library(readr)
#training data 
moneyball_training_data <- read_csv("moneyball-training-data.csv")
View(moneyball_training_data)
#evaluation data 
moneyball_evaluation_data <- read_csv("moneyball-evaluation-data.csv")
View(moneyball_evaluation_data)
```
For our moneyball training data set we have 2276 observations and 17 columns. For easy exploration, let's describe the group of variables first, and exclude INDEX since we will not need it . Please see as follows:

TARGET_WINS - our dependent variable

TEAM_BATTING - refers to a team's offensive ability when hitting the ball

TEAM_BASERUN - determines how efficiently a team navigates the bases after hitting the ball

TEAM_FIELDING - is the defensive skill of catching and throwing the ball to record outs

TEAM_PITCHING - represents the ability of a team's pitchers to throw strikes and prevent the opposing team from hitting the ball effectively
```{r setup1, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
#remove index variable in our dataset
moneyball_training_data = moneyball_training_data[,-1]

# check summary statistic 
summary(moneyball_training_data)
```
The average number of team wins is approximately 81, with a range from 0 to 146, suggesting a wide variance in the teams' success. The team typically has 1,469 hits per season, with notable variability in extra-base hits such as doubles (mean = 241) and triples (mean = 55). Home runs range from 0 to 264, while walks and strikeouts average around 502 and 736, respectively. Pitching statistics reveal that teams allow an average of 1,779 hits and 105 home runs, while fielding metrics indicate an average of 246 errors per season. There is also the presence of missing values which could be attributed to many things that can occur during a game.


To help us further with our exploration, let's have quick overview of our histogram and q-q plot to check for the normality of distribution of our data and to easily spot outliers that might suggest to fix our missing values or transform our variables later.

```{r setup2, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
# let's plot a histogram using DataExplorer library
plot_histogram(moneyball_training_data)

# let's plot a qq plot
plot_qq(moneyball_training_data)
```

With an overview of both graphs, we can see that only some of the variables are normally distributed, let's say the variables TEAM_BATTING_2B, TEAM_BATTING_HBP, and TEAM_FIELDING_DP, the others are skewed to the right with reference to our histogram, either having a curve or a heavy tail with reference to our q-q plot.

Our goal is to predict the team's target wins, so analyzing correlations will help identify which variables have the strongest impact on winning outcomes.
```{r setup3, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
cor_matrix <- cor(moneyball_training_data, use = "complete.obs")
cor_matrix["TARGET_WINS", ]

```
The strongest positive correlations with wins are seen in TEAM_BATTING_H (+0.35), TEAM_BATTING_BB (+0.30), and TEAM_BATTING_2B (+0.26), indicating that hits, walks, and doubles contribute significantly to scoring runs. Conversely, TEAM_FIELDING_E (-0.17) and TEAM_PITCHING_H (-0.08) show negative correlations, suggesting that poor defense and allowing more hits can reduce wins. Weak correlations, such as TEAM_BATTING_SO (-0.03) and TEAM_FIELDING_DP (-0.03), indicate that strikeouts and double plays have little impact. For model building, TEAM_BATTING_H, TEAM_BATTING_BB, TEAM_BATTING_2B, TEAM_BATTING_HR, and TEAM_PITCHING_HR should be prioritized as key predictive features.

DATA PREPARATION
Since we know that there are missing variables that can affect the distribution, we will first identify and address them. Then, we will clean the data to improve normality in the distribution.
```{r setup4, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
#How many missing variables ?
missing <- colSums(moneyball_training_data %>% sapply(is.na))
missing_pct <- round(missing / nrow(moneyball_training_data) * 100, 2)
stack(sort(missing_pct, decreasing = TRUE))
```
TEAM_BATTING_HBP has 91.61% missing values, so the best approach is to remove this variable entirely. TEAM_BASERUN_CS (33.92%) and TEAM_FIELDING_DP (12.57%) also have missing data, but removing them may not be ideal. Instead, we can use imputation techniques like using mean and median to fill the missing variables and then doing a log/square root transformation because the data is showed being skewed right and it would help the distribtion be more normal. 

```{r setup5, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

#remove Batting
moneyball_training_data <- moneyball_training_data %>% select(-TEAM_BATTING_HBP)

# using mean and median to impute the missing variables
moneyball_training_data$TEAM_BASERUN_CS[is.na(moneyball_training_data$TEAM_BASERUN_CS)] <- 
  median(moneyball_training_data$TEAM_BASERUN_CS, na.rm = TRUE)

moneyball_training_data$TEAM_FIELDING_DP[is.na(moneyball_training_data$TEAM_FIELDING_DP)] <- 
  median(moneyball_training_data$TEAM_FIELDING_DP, na.rm = TRUE)

moneyball_training_data$TEAM_BATTING_SO[is.na(moneyball_training_data$TEAM_BATTING_SO)] <- 
  median(moneyball_training_data$TEAM_BATTING_SO, na.rm = TRUE)

# log transformation to reduce skewness
skewed_vars <- c("TEAM_BASERUN_CS", "TEAM_FIELDING_DP", "TEAM_FIELDING_E", 
                 "TEAM_PITCHING_SO", "TEAM_PITCHING_BB")

moneyball_training_data[skewed_vars] <- lapply(moneyball_training_data[skewed_vars], log1p)

# check summary statistic to see if there is a difference 
summary(moneyball_training_data)
```

After applying log transformation and missing value imputation, several key variables changed. TEAM_BASERUN_CS and TEAM_FIELDING_DP now have smaller ranges, confirming that log transformation compressed their values. TEAM_PITCHING_BB and TEAM_FIELDING_E also show reduced variability, meaning extreme outliers have less influence. Additionally, missing values in TEAM_BASERUN_CS and TEAM_FIELDING_DP were replaced using the median, making these variables more complete.


BUILD MODELS 

MODEL ONE - SIMPLE LINEAR REGRESSION
For Model 1, a simple linear regression was used to predict TARGET_WINS based on TEAM_BATTING_H (hits), TEAM_BATTING_HR (home runs), and TEAM_BATTING_BB (walks). These variables were chosen because they represent key offensive metrics that contribute to scoring runs and, ultimately, winning games.
```{r setup5, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

model1 <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_BB, data = moneyball_training_data)
summary(model1)
```
This regression model predicts TARGET_WINS using TEAM_BATTING_H (hits), TEAM_BATTING_HR (home runs), and TEAM_BATTING_BB (walks). The results show that hits have the strongest positive effect on wins, where every additional hit increases wins by 0.044 (p < 2e-16). This is followed by walks, which increase wins by 0.029 (p < 2e-16), and home runs, which increase wins by 0.015 (p = 0.0054). The model is statistically significant (p < 2.2e-16), but with an R² of 0.2221, it only explains 22.2% of the variance in wins, suggesting that additional factors such as pitching and defense may improve prediction accuracy. The residual standard error of 13.9 indicates that there is still considerable variability in the model’s predictions.
