---
title: "Crimes"
author: "Bedros, Haig; Dela Cruz, Sandra; Hugh, Tiffany; Li, Yanyi"
date: "2025-03-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(DataExplorer)
#library(correlationfunnel)
library(reshape2)
library(pROC)
library(readr)
```

```{r, include=FALSE}
# Import dataset
crime_training_data <- read_csv("https://raw.githubusercontent.com/tiffhugh/DATA-621/refs/heads/main/Assignment%20%233/crime-training-data_modified.csv")
View(crime_training_data)

crime_evaluation_data<- read_csv("https://raw.githubusercontent.com/tiffhugh/DATA-621/refs/heads/main/Assignment%20%233/crime-evaluation-data_modified.csv")
View(crime_evaluation_data)
```

### Objective 
For our third group assignment, we were provided with a dataset containing crime-related information for various neighborhoods in a major city. The dataset includes a binary response variable indicating whether a neighborhood’s crime rate is above the median (1) or below it (0). Our goal is to develop a binary logistic regression model using the training dataset to predict whether a given neighborhood is at risk for high crime levels. Through this analysis, we aim to identify key predictors of crime and assess the model's effectiveness in classification.

The variables in this data compromise of: 
• zn: proportion of residential land zoned for large lots (over 25000 square feet) (predictor variable)

• indus: proportion of non-retail business acres per suburb (predictor variable)

• chas: a dummy var. for whether the suburb borders the Charles River (1) or not (0) (predictor variable)

• nox: nitrogen oxides concentration (parts per 10 million) (predictor variable)

• rm: average number of rooms per dwelling (predictor variable)

• age: proportion of owner-occupied units built prior to 1940 (predictor variable)

• dis: weighted mean of distances to five Boston employment centers (predictor variable)

• rad: index of accessibility to radial highways (predictor variable)

• tax: full-value property-tax rate per $10,000 (predictor variable)

• ptratio: pupil-teacher ratio by town (predictor variable)

• lstat: lower status of the population (percent) (predictor variable)

• medv: median value of owner-occupied homes in $1000s (predictor variable)

• target: whether the crime rate is above the median crime rate (1) or not (0) (response variable) 


# **Data Exploration** 

```{r, echo=FALSE}
cat("Summary of training data: \n")
summary(crime_training_data)

cat("Summary of evaluation data: \n")
summary(crime_evaluation_data)
```

Preliminary summary statistics reveal that the crime training dataset consists of 466 observations with 12 predictor variables, while the crime evaluation dataset contains 40 observations with 13 variables, including the target variable. Our next step is to examine the distribution of the predictor variables to gain deeper insights into their characteristics and potential impact on the model.

### Histogram of Predictor Variables
```{r echo=FALSE}
#| fig-height: 5
#| fig-width: 10
# let's plot a histogram using DataExplorer library
plot_histogram(crime_training_data)
```

The histograms reveal distinct distribution patterns among the variables, shedding light on their underlying characteristics. The zn variable, representing the proportion of land zoned for large lots, is heavily right-skewed, indicating that most suburbs have little to no large-lot zoning, with only a few having substantial allocations. The indus variable follows a moderately right-skewed distribution, suggesting that non-retail business acreage is relatively low in most suburbs but significantly higher in a few areas. The chas variable, being binary (0 or 1), is highly imbalanced, with most suburbs not bordering the Charles River. The nox variable, which measures air pollution, has a somewhat uniform but slightly skewed distribution, reflecting variability in nitrogen oxide concentrations across different areas.

The rm variable, capturing the average number of rooms per dwelling, is approximately normal, implying that most homes have a standard number of rooms, with fewer extreme cases. The age variable is skewed toward higher values, highlighting that a significant proportion of owner-occupied units were built before 1940. The dis variable is right-skewed, meaning that while some suburbs are close to employment centers, many are farther away. The rad and tax variables exhibit a bimodal distribution, which makes sense given their categorical-like nature—certain suburbs have particularly high accessibility to highways or high property tax rates, while others do not.

The ptratio variable is also right-skewed, showing that most suburbs have lower pupil-teacher ratios, with a few having significantly higher ones. The lstat variable, representing the percentage of lower-status population, has a broader spread, signifying substantial variation in socioeconomic conditions across suburbs. Finally, the medv variable, indicating the median home value, follows a near-normal distribution but with slight skewness, suggesting that while home values are mostly centered around a typical range, some suburbs have notably high property values. These distribution characteristics highlight potential areas for transformation or deeper statistical exploration.

# **Data Preparation**

### Missing Data

```{r, echo=FALSE}
#| fig-height: 3
#| fig-width: 8
plot_missing(crime_training_data)
```

As we can see in the graph above that there is no missing data with our any
of our predictor variables.

### Outliers

```{r, echo=FALSE}
#| fig-height: 5
#| fig-width: 10
plot_boxplot(crime_training_data, by='target')
```
The boxplot above separates the data by the 'target' variable, showing the distribution of crime rates for the two target groups (0 and 1). The median line inside each box indicates the central value of the crime rate for each group. From the boxplot, we can observe that the distributions of crime rates for both target = 0 (box below) and target = 1 (box above) are quite widely spread. This suggests that there is significant variability in the crime rate for both groups, with several outliers beyond the whiskers. Given the broad distribution in both groups, we choose to retain the outliers in our analysis, as they may represent important extreme events, such as rare high-crime instances. 

### Combine Variables

Older homes might be located in neighborhoods with lower home values, and this combination could provide insight into areas that may be socioeconomically distressed and more prone to higher crime rates. Therefore, we'll introduce a new variable called "housing_age_value_ratio." This variable measures the ratio of the proportion of owner-occupied units built prior to 1940 (age) divided by the median value of owner-occupied homes in thousands of dollars (medv).

```{r, echo=FALSE}
crime_training_data$housing_age_value_ratio <- 
  crime_training_data$age / crime_training_data$medv
head(crime_training_data[, c("age", "medv", "housing_age_value_ratio")])
```

### Correlation Funnel
```{r, echo=FALSE}
# Step 1: Convert to Binary Format
crime_training_data_binarized_tbl <- crime_training_data %>% 
  binarize(n_bins = 4, thresh_infreq = 0.01)

# Step 2: Perform Correlation Analysis
crime_training_data_correlated_tbl <- crime_training_data_binarized_tbl %>% 
  correlate(target = target__1)

head(crime_training_data_correlated_tbl)
```

```{r, echo=FALSE}
#| fig-height: 5
#| fig-width: 10
# Step 3: Visualize the Correlation Funnel
crime_training_data_correlated_tbl %>%
   plot_correlation_funnel(interactive = FALSE)
```
By using the correlation funnel above, we can guide our decision on which variables to choose when building our model. Our target variable for this funnel is target = 1, which indicates that the crime rate is above the median crime rate. The variables at the top of the funnel have the highest correlation with the target, and as we move down the funnel, the correlation becomes weaker.

# **Build Models**

### Model #1 : General Linear Models


```{r, echo=TRUE}
# build glm
logit_model_1 <- glm(target ~ ., 
                     data = crime_training_data, 
                     family = binomial)

# summarize the model
summary(logit_model_1)
```

The first model selected a general linear model to observe the relationships between all the predictors and the target variable. The significant predictors include nox, dis, rad, tax, ptratio, and medv:

• Nitrogen oxides concentration: The coefficient is 49.34, the largest positive coefficient, indicating that higher levels of nitrogen oxides are strongly associated with higher crime rates. This is statistically significant with a p-value of 4.61e-10, making it a strong predictor. This result is intuitive, as areas with higher pollution levels could indicate lower socioeconomic conditions, which may correlate with higher crime rates.

• Distance to employment centers: The coefficient is 0.73, suggesting that a higher distance to employment centers increases the odds of a neighborhood having a high crime rate. With a p-value of 0.00160, this relationship is statistically significant.

• Accessibility to radial highways: The coefficient is 0.71, meaning that higher accessibility to radial highways increases the likelihood of a high crime rate. This is statistically significant with a p-value of 3.64e-05. While it's not entirely surprising that better highway access could correlate with higher crime (since highways can facilitate criminal activity), it’s still a bit counterintuitive because areas with better infrastructure might also be expected to have more resources that could reduce crime.

• Property tax rate: The coefficient is -0.0073, indicating that a higher property tax rate slightly decreases the odds of a neighborhood having a high crime rate. This is statistically significant with a p-value of 0.02249. This is counterintuitive because one might expect higher property taxes to correlate with higher crime in economically disadvantaged areas. However, it could be related to gentrification, where rising property taxes are often associated with neighborhood revitalization, improved infrastructure, and a decrease in crime.

• Pupil-teacher ratio: The coefficient is 0.39, suggesting that a higher pupil-teacher ratio increases the odds of a neighborhood having a high crime rate. With a p-value of 0.00186, this relationship is statistically significant.

• (Median value of owner-occupied homes: The coefficient is 0.21, suggesting that higher home values are associated with higher crime rates. This is statistically significant with a p-value of 0.00475.

Other predictors, such as zn, indus, chas, rm, age, lstat, and housing_age_value_ratio, are not significant predictors:

• Housing_age_value_ratio: The coefficient is 0.27, indicating a slight positive association with high crime rates, but the p-value of 0.28631 suggests it is not statistically significant.

• Dummy for bordering the Charles River: The coefficient is 0.95, indicating that neighborhoods bordering the Charles River have higher odds of having a high crime rate. However, the p-value of 0.21422 suggests this relationship is not statistically significant.

• Proportion of non-retail business acres: The coefficient is -0.06, meaning that a one-unit increase in the proportion of non-retail business acres decreases the log-odds of a high crime rate by approximately 0.063. The p-value of 0.18764 indicates that this predictor is not statistically significant and might not be useful in predicting crime rates.

This analysis suggests that while some predictors significantly influence crime rates, others do not show strong associations, and their inclusion in the model could be reconsidered. A transformation can be done to see how more align predictors are to the target. 
