---
title: "HW5"
author: "Bedros, Haig; Dela Cruz, Sandra; Hugh, Tiffany; Li, Yanyi"
date: "`r Sys.Date()`"
output:
editor_options: 
  markdown: 
    wrap: 72
---
<style type="text/css">

h1.title {
  font-size: 26px;
  color: Black;
  text-align: center;
}
h4.author { 
  font-size: 18px;
  color: Black;
  text-align: center;
}
h4.date {
  font-size: 12px;
  color: Black;
  text-align: center;
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 50), tidy = TRUE, warning = FALSE)
```

```{r, include=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)
library(purrr)
library(DataExplorer)
library(reshape2)
library(pROC)
library(readr)
library(caret)
library(DHARMa)
library(lares)
library(mice)
```

```{r, include=FALSE}
# Import dataset
wine_training_data <- read_csv("https://raw.githubusercontent.com/tiffhugh/DATA-621/refs/heads/main/Assignment%20%235/wine-training-data.csv")
wine_evaluation_data<- read_csv("https://raw.githubusercontent.com/tiffhugh/DATA-621/refs/heads/main/Assignment%20%235/wine-evaluation-data.csv")
```

# **Data Exploration**
First, let's remove the 'INDEX' column in both of the dataset, since we are not going to need it.
```{r, echo=FALSE}
# remove index columns
wine_training_data <- wine_training_data %>% select(-INDEX)
glimpse(wine_training_data)
dim(wine_training_data)

# do the same for evaluation data
wine_evaluation_data <- wine_evaluation_data %>% select(-IN)
```
Our wine training dataset has 12,795 observations and 15 columns. As we can see, there are a lot of data transformations that needs to be done with our variables, like cleaning our values and changing the columns into categorical values, like for `LabelAppeal` and `STARS`.


