---
title: "HW4"
author: "Bedros, Haig; Dela Cruz, Sandra; Hugh, Tiffany; Li, Yanyi"
date: "`r Sys.Date()`"
output: 
editor_options: 
  markdown: 
    wrap: 72
---
<style type="text/css">

h1.title {
  font-size: 26px;
  color: Black;
  text-align: center;
}
h4.author { 
  font-size: 18px;
  color: Black;
  text-align: center;
}
h4.date {
  font-size: 12px;
  color: Black;
  text-align: center;
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 50), tidy = TRUE, warning = FALSE)
```

```{r, include=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)
library(purrr)
library(DataExplorer)
library(reshape2)
library(pROC)
library(readr)
library(caret)
library(DHARMa)
library(lares)
library(mice)
```

```{r, include=FALSE}
# Import dataset
insurance_training_data <- read_csv("https://raw.githubusercontent.com/tiffhugh/DATA-621/refs/heads/main/Assignment%20%234/insurance_training_data.csv")
insurance_evaluation_data<- read_csv("https://raw.githubusercontent.com/tiffhugh/DATA-621/refs/heads/main/Assignment%20%234/insurance-evaluation-data%20(1).csv")
```

# **Data Exploration**

First, let's remove the 'INDEX' column in both of the dataset, since we are not going to need it.
```{r, echo=FALSE}
# remove index columns
insurance_training_data <- insurance_training_data %>% select(-INDEX)
glimpse(insurance_training_data)
dim(insurance_training_data)

# do the same for evaluation data
insurance_evaluation_data <- insurance_evaluation_data %>% select(-INDEX)
```
Our insurance training dataset has 8,161 observations and 25 columns. As we can see, there are a lot of data transformations that needs to be done with our variables, like cleaning our values and changing the columns into categorical or numerical.

### Data Transformation of Variables
First, let's clean our values by removing unecessary characters such as 'z_', '$', ',', and '-'.
```{r, echo=FALSE}
# clean the values in our variables by removing 'z_', '$', and ','
insurance_training_data <- insurance_training_data %>% 
  mutate(across(everything(), ~gsub("z_|\\$|\\,|\\-","", .x)))
glimpse(insurance_training_data)

insurance_evaluation_data <- insurance_evaluation_data %>%
  mutate(across(everything(), ~gsub("z_|\\$|\\,|\\-", "", .x)))
```
Now that we have removed these unnecessary characters in our values, let's review our distinct values.
```{r, echo=FALSE}
# check distinct values for all columns
insurance_training_data %>% summarise(across(everything(), ~n_distinct(.)))

# check unique values in possibly categorical columns
insurance_training_data %>% 
  distinct(KIDSDRIV, PARENT1, MSTATUS, SEX, EDUCATION, JOB, CAR_USE, CAR_TYPE,
           RED_CAR, REVOKED, URBANICITY) %>% 
  map(~unique(.))
```
As we can see, 'NA' is present for the `JOB` column. Further investigation with our dataset showed that these individuals with 'NA' in the `JOB` column completed either 'Masters' or 'PhD'; however, we'll fill in the 'NA' with 'Other' just for general consideration and also to prevent bias. 
```{r, echo=FALSE}
# fill the missing 'JOB' values with 'Self-employed/ Freelance'
insurance_training_data <- insurance_training_data %>% 
  mutate(JOB = ifelse(is.na(JOB), 'Other', JOB))
# view changes
insurance_training_data %>% 
  distinct(JOB)

# do the same for evaluation data
insurance_evaluation_data <- insurance_evaluation_data %>% 
  mutate(JOB = ifelse(is.na(JOB), 'Other', JOB))
```
Now, let's proceed with converting our columns to the appropriate data types, such as factors, numeric, or ordinal variables
Convert to factor variables: `TARGET_FLAG`, `KIDSDRIV`, `MSTATUS`, `SEX`, `EDUCATION`, `JOB`, `CAR_USE`, `CAR_TYPE`, `RED_CAR`, `REVOKED`, `URBANICITY`
Convert to numeric variables: `TARGET_AMT`, `AGE`, `HOMEKIDS`, `YOJ`, `INCOME`, `HOME_VAL`, `TRAVTIME`, `BLUEBOOK`, `TIF`, `OLDCLAIM`, `CAR_AGE`, `URBANICITY`
Convert to ordinal variables: `CLM_FREQ`, `MVR_PTS`
```{r, echo=FALSE}
insurance_training_data <- insurance_training_data %>% 
  mutate(
    # convert all factor variables
    across(c(TARGET_FLAG, KIDSDRIV, MSTATUS, SEX, EDUCATION, JOB, CAR_USE,
             CAR_TYPE, RED_CAR, REVOKED, URBANICITY), as.factor),
    # convert all numeric variables
    across(c(TARGET_AMT, AGE, HOMEKIDS, YOJ, INCOME, HOME_VAL, TRAVTIME,
             BLUEBOOK, TIF, OLDCLAIM, CAR_AGE, URBANICITY), as.numeric),
    # convert all ordinal variables
    across(c(CLM_FREQ, MVR_PTS), as.ordered)
  )

glimpse(insurance_training_data)

# do the same for evaluation data
insurance_evaluation_data <- insurance_evaluation_data %>% 
  mutate(
    # convert all factor variables
    across(c(TARGET_FLAG, KIDSDRIV, MSTATUS, SEX, EDUCATION, JOB, CAR_USE,
             CAR_TYPE, RED_CAR, REVOKED, URBANICITY), as.factor),
    # convert all numeric variables
    across(c(TARGET_AMT, AGE, HOMEKIDS, YOJ, INCOME, HOME_VAL, TRAVTIME,
             BLUEBOOK, TIF, OLDCLAIM, CAR_AGE, URBANICITY), as.numeric),
    # convert all ordinal variables
    across(c(CLM_FREQ, MVR_PTS), as.ordered)
  )
```
### Summary of Datasets
Here we see 'NA' values with `AGE`, `YOJ`, `INCOME`, `HOME_VAL`, and `CAR_AGE`. Later we'll impute each of these missing values.

```{r, echo=FALSE}
#| fig-height: 3
#| fig-width: 8
cat("Summary of training data: \n")
summary(insurance_training_data)
```
### Histogram of Continuous Variables
```{r, echo=FALSE}
#| fig-height: 5
#| fig-width: 10
plot_histogram(insurance_training_data)
plot_boxplot(insurance_training_data, by='TARGET_FLAG')
```
With the view of our histogram, we can see that most of our continuous data are skewed to the right.

### Bar Chart of Categorical Variables
```{r, echo=FALSE}
#| fig-height: 5
#| fig-width: 10
plot_bar(insurance_training_data)
```
Our bar chart reveals a lot about our dataset, such as the fact that most of our customers have 'Blue Collar' jobs, own either an 'SUV' or 'Minivan', are 'High School Graduates' or hold a 'Bachelor's' degree, are single parents (based on `PARENT1`), do not have a ticket (based on `MVR_PTS`), and live in urban areas.

### Boxplot of Some Variables
```{r, echo=FALSE}
#| fig-height: 5
#| fig-width: 10
plot_boxplot(insurance_training_data, by='TARGET_FLAG')
```
The boxplot above separates the data by the TARGET_FLAG variable. Although there are noticeable outliers in our data, we will retain them as they make sense in real-world scenarios. For example, someone could have worked for a long time at a particular company (variable `YOJ`), or someone might use their car for commercial purposes (variable `TRAV_TIME`).

### Visual Correlation of Variables
```{r, echo=FALSE, message=FALSE}
#| fig-height: 5
#| fig-width: 10
corr_cross(insurance_training_data, 
           method = "spearman", max_pvalue = 0.05, top = 20)

corr_cross(insurance_training_data, 
           method = "spearman", max_pvalue = 0.05, 
           contains = c('TARGET_AMT', 'TARGET_FLAG'))
```
To visualize correlations, we use the 'corr_cross' function from the 'lares' package, which allows us to easily compare variables. We set the method to 'Spearman' since we know that most of our data is not normally distributed. In the first cross-correlation graph, we can see that 'OLDCLAIM' is highly correlated with 'CLM_FREQ,' with correlations gradually decreasing as we move down the list to variables with weaker correlations. In the second correlation graph, we observe that our predictor variables are not strongly correlated with the target variables. With all this information, we can now proceed with our data preparation.

# **Data Preparation**
### Change of Variable
For the variable `HOME_VAL`, we observed that there are 2,294 instances where the value is zero. We have made the assumption that these individuals are likely renters, as they do not own a home. For the records where `HOME_VAL` is missing, we have decided to label them as 'Unknown'. This is because we are unsure of their homeownership status, and by assigning them as 'Unknown', we avoid introducing any bias into our analysis.

```{r, echo=FALSE}
# Count zero's in 'HOME_VAL'
zero_count <- sum(insurance_training_data$HOME_VAL == 0, na.rm = TRUE)
cat("Zero count for HOME_VAL: \n")
print(zero_count)
```

In this step, we are creating a new variable called `HOME_STATUS` based on the values of `HOME_VAL`:

- Individuals with`HOME_VAL` > 0 will be categorized as 'Homeowners'.

- Individuals with `HOME_VAL` == 0 will be categorized as 'Renters'.

- Missing values (NA) in `HOME_VAL` will be labeled as 'Unknown', as we cannot determine their homeownership status.
```{r, echo=FALSE}
# create new variable 'HOME_STATUS' to separate renters, homeowners, and unknown
insurance_training_data <- insurance_training_data %>% 
  mutate(HOME_STATUS = case_when(
    is.na(HOME_VAL) ~ "Unknown",
    HOME_VAL == 0 ~"Renter",
    TRUE ~ "Homeowner"
  )) %>% 
  mutate(HOME_STATUS = factor(HOME_STATUS, levels = c("Homeowner", "Renter", "Unknown")))

head(insurance_training_data[, c("HOME_VAL", "HOME_STATUS")])

# do the same for evaluation dataset
insurance_evaluation_data <- insurance_evaluation_data %>% 
  mutate(HOME_STATUS = case_when(
    is.na(HOME_VAL) ~ "Unknown",
    HOME_VAL == 0 ~"Renter",
    TRUE ~ "Homeowner"
  )) %>% 
  mutate(HOME_STATUS = factor(HOME_STATUS, levels = c("Homeowner", "Renter", "Unknown")))
```

Now that we categorize the homeownership, let's delete the `HOME_VAL` column for us to proceed in other imputation of our columns.
```{r, echo=FALSE}
insurance_training_data <- insurance_training_data %>% 
  select(-HOME_VAL)
colnames(insurance_training_data)

insurance_evaluation_data <- insurance_evaluation_data %>% 
  select(-HOME_VAL)
```
### Imputation of Variables
For the variables `AGE`, `YOJ`, `INCOME`, and `CAR_AGE`, we will use Predictive Mean Matching (PMM) imputation from the mice package. PMM is a general imputation method that ensures missing values are replaced with observed values, helping to maintain the original relationships in the data, including non-linear ones. We’ll set the number of imputations (m) to 5, as it’s a large dataset and this is a standard choice. We will also leave the default maximum iterations (maxit) at 5."
```{r, echo=FALSE}
# mice imputation
input_training = insurance_training_data
method_vector <- c("", "", "", "pmm", "",
                   "pmm", "pmm", "", "", "",
                   "", "", "", "", "",
                   "", "", "", "", "",
                   "", "", "pmm", "", ""
                   )
mice_imp = mice(input_training, m = 5, 
                method = method_vector)
summary(input_training$AGE)
mice_imp$imp$AGE
clean_insurance_training_data <- complete(mice_imp, 4)
summary(clean_insurance_training_data)
plot_histogram(clean_insurance_training_data)
```
